---
title: |
       | Supporting Information 2: Exploring correlations in genetic and cultural variation across   language families in Northeast Asia 
       |
       | Preprocessing: Grammar and Phonology^[The R code for this analysis was written by Balthasar Bickel, Peter Ranacher, and Damián E. Blasi]
output: 
  pdf_document:
    keep_tex: yes
    latex_engine: xelatex
    fig_crop: true
bibliography: ../genetics.bib
header-includes:
 - \usepackage[Symbol]{upgreek}
 - \usepackage{tocloft}
 - \settowidth{\cftsubsecnumwidth}{S4.5x}
 - \settowidth{\cftsecnumwidth}{S4.i}
 - \setcounter{table}{12}
 - \renewcommand{\thetable}{S\arabic{table}}
---


```{r setup, include=FALSE}
library(knitr) # A General-Purpose Package for Dynamic Report Generation in R
library(kableExtra) # Construct Complex Table with 'kable' and Pipe Syntax, CRAN v1.3.1
library(tidyverse) # Easily Install and Load the 'Tidyverse', CRAN v1.3.0

opts_chunk$set(fig.path = 'figures/',
           dev = 'cairo_pdf', dev.args = list(bg = 'transparent'), 
                fig.height = 7,
                fig.width = 14,
                message = F,
                warning = F,
                autodep = T,
                cache.comments = F,
                crop = T,
                pars = T,
                out.extra = ''
                )
options(width = 180, knitr.kable.NA = '', knitr.table.format = "latex")
```

Data on grammar and phonology are aggregated from the following sources:

- AUTOTYP (*34*)
- WALS (*33*), enriched by recodings (*65*)
- ANU Phonotactics database (*35*)
- PHOIBLE (*36*)

The split into phonology vs. grammar is based on the broad definition in `categorization_of_variables.csv` which includes phonotactic and morphophonological data under phonology.^[Available at  https://github.com/derpetermann/music_languages_genes] We extract a subset from the data comprising languages from fourteen different societies. Note that in AUTOTYP and WALS, Buriat is represented by ISO code `bua` (Buriat in general), while in PHOIBLE  it is represented by ISO code  `bxr` and in the ANU data by `bxm`. We mapped all of them below to `bua`. For details, see Supporting Information: Variables for Grammar and Phonology. 
```{r lingdata, cache=T}

# Define single languages and family-related languages
single_lgs <- c("Ainu", "Buryat", "Japanese", "Korean", "Nivkh", "West Greenlandic", 
                "Yakut", "Yukagir")
uralic <- c("Selkup", "Nganasan")
chkkat <- c("Chukchi", "Koryak")
tungus <- c("Even", "Evenki")

# Read categorization of variables
var_categories <- read.csv("../data/typology/categorization_of_variables.csv",
                           stringsAsFactors = F)


# Read the grammar, phonology and typlogy data (meta data)
typology.list <- readRDS("../data/typology/typology.list.RDS")
phonological_vars <- unlist(as.character(sapply(typology.list, function(x) {
        var_categories[var_categories$Variable == x[1, "variable.ID"], 
                       "Broad_Phon_Definition_Binary"] == "Phonology" }) ))
phonology_list <- typology.list[phonological_vars == "TRUE"]
grammar_vars <- unlist(as.character(sapply(typology.list, function(x) {
        var_categories[var_categories$Variable == x[1, "variable.ID"], 
                       "Broad_Phon_Definition_Binary"] == "Grammar" })))
grammar_list <- typology.list[grammar_vars == "TRUE"]
typology_coverage_df <- readRDS("../data/typology/typology.coverage.RDS")

# Define subset
siberia_sample <- c(
 "[i-ain][a-12][g-ainu1240]",    # Ainu
 "[i-bua][a-1095][g-buri1258]",  # Buriat
 "[i-bxm][a-][g-mong1330]",      # Buriat (Mongolia)
 "[i-bxr][a-][g-russ1264]",      # Buriat (Russia)
 "[i-ckt][a-56][g-chuk1273]",    # Chukchi
 "[i-eve][a-738][g-even1260]",   # Even
 "[i-evn][a-527][g-even1259]",   # Evenki
 "[i-kal][a-511][g-kala1399]",   # West Greenlandic
 "[i-jpn][a-118][g-nucl1643]",   # Japanese
 "[i-kor][a-141][g-kore1280]",   # Korean
 "[i-kpy][a-1808][g-kory1246]",  # Koryak
 "[i-nio][a-2172][g-ngan1291]",  # Nganasan
 "[i-niv][a-433][g-gily1242]",   # Nivkh
 "[i-sel][a-2393][g-selk1253]",  # Selkup
 "[i-sah][a-2662][g-yaku1245]",  # Yakut
 "[i-ykg][a-423][g-nort2745]")   # Yukagir (Tundra)

# Extract meta data for the above subset
siberia_metadata_all <- subset(typology_coverage_df, UULID %in% siberia_sample)
siberia_metadata <- subset(siberia_metadata_all, !isocode %in% c('bxm','bxr'))

counts <- xtabs(~autotyp.Stock, siberia_metadata, drop.unused.levels = T)
rownames(siberia_metadata) <- with(siberia_metadata,
                                   ifelse(autotyp.Stock %in% names(counts[counts>1]),
                                          paste(autotyp.Stock, autotyp.Language, sep="/"),
                                          paste(autotyp.Language)))

rownames(siberia_metadata) <- gsub('Yukagir/','', rownames(siberia_metadata))
```


We only use variables with one data point per language, and only variables with non-constant values (which otherwise can't deliver a distance signal). At the same time, we also remap `bxr` and `bxm` to `bua` (cf. above). We trim the linguistic data accordingly.

```{r trimllingdata, cache=T}

trim_data <- function(data.list, trim.to=siberia_metadata_all, extra.coverage=.8) {
 #' Trims the linguistic input data and only keeps variables with
 #' data points for languages in the study area and with non-constant values

 #' @param data.list the data to be trimmed
 #' @param trim.to contains the ids of those languages that are retained after trimming  
 #' @param extra.coverage the percentage of covered data
 #' @return the trimmed data

  lgs <- lapply(data.list, function(l) {
	  l$UULID <- ifelse(l$isocode %in% c('bxm','bxr'),
	  		   '[i-bua][a-1095][g-buri1258]',
		      paste(l$UULID))
	  subset(l, UULID %in% trim.to$UULID)
	 })
  vars <- lgs[sapply(lgs, function(l) {
	   length(l$UULID)==length(unique(l$UULID)) &
	   length(unique(l[,1]))>1 &
	     length(unique(l$UULID)) >= floor(extra.coverage*length(trim.to$UULID))
	  })]
  return(lapply(vars, function(l) l[,c(1,3)]))}

# Trim the data
siberia_grammar_list <- trim_data(grammar_list,extra.coverage = 0.8)
siberia_phonology_list <- trim_data(phonology_list,extra.coverage = 0.8)
```

The following lists the phonological variables we captured. For full definitions and descriptions of the variables, see the source databases listed above.
```{r phonvar, echo=F, cache=T}
cat(paste('-', names(siberia_phonology_list)), sep='\n')
```

And these the grammar variables:

```{r grammvar, echo=F, cache=T}
cat(paste("-", names(siberia_grammar_list)), sep = "\n")
```

For each society we compute the coverage, i.e. the percentage of available variables per society.  

```{r lingcoverage, message=FALSE, warning=FALSE, cache=T}

compute_coverage <- function(data.list, gg=siberia_metadata) {
 #' Computes the coverage of all languages
 #' @param data.list the input data 
 #' @param gg a data.frame comprising the languages for which the coverage is computed
 #' @return the input data with the coverage added as a separate column

  x <- sapply(gg$UULID, function(l) {
          coverage <- round(mean(sapply(data.list, function(v) { l %in% v$UULID })),2)*100
        })
  df <- data.frame(UULID=names(x), Coverage=x)
  gg$Language <- rownames(gg)
  df.g <- merge(df, gg)
  return(df.g)}

# Compute the coverage for each variable
grammar_coverage <- compute_coverage(siberia_grammar_list) %>% 
  dplyr::select(Language, Coverage)
phonology_coverage <- compute_coverage(siberia_phonology_list) %>% 
  dplyr::select(Language, Coverage)
```

We simplify and standardize the language names and visualize the coverage in a table. Finally, we flatten the nested linguistic data and convert them to data frames. 
```{r lingcleanup, message=FALSE, warning=FALSE, cache=T, echo=F}

# Change variable names (see next section)
# Grammar
grammar_coverage[grammar_coverage$Language == 'Chukchi-Kamchatkan/Chukchi', 
                 "Language"] <-'Chukchi'
grammar_coverage[grammar_coverage$Language == 'Tungusic/Evenki', "Language"] <- 'Evenki'
grammar_coverage[grammar_coverage$Language == 'Greenlandic Eskimo (West)', 
                 "Language"] <-'West Greenlandic'
grammar_coverage[grammar_coverage$Language == 'Uralic/Selkup', "Language"] <- 'Selkup'
grammar_coverage[grammar_coverage$Language == 'Yukagir (Tundra)', "Language"] <- 'Yukagir'
grammar_coverage[grammar_coverage$Language == 'Tungusic/Even', "Language"] <- 'Even'
grammar_coverage[grammar_coverage$Language == 'Buriat', "Language"] <- 'Buryat'
grammar_coverage[grammar_coverage$Language == 'Uralic/Nganasan', "Language"] <- 'Nganasan'
grammar_coverage[grammar_coverage$Language == 'Chukchi-Kamchatkan/Koryak', 
                 "Language"] <- 'Koryak'

# Phonology
phonology_coverage[phonology_coverage$Language == 'Chukchi-Kamchatkan/Chukchi', 
                   "Language"] <-'Chukchi'
phonology_coverage[phonology_coverage$Language == 'Tungusic/Evenki', "Language"] <- 'Evenki'
phonology_coverage[phonology_coverage$Language == 'Greenlandic Eskimo (West)', 
                   "Language"] <-'West Greenlandic'
phonology_coverage[phonology_coverage$Language == 'Uralic/Selkup', "Language"] <- 'Selkup'
phonology_coverage[phonology_coverage$Language == 'Yukagir (Tundra)', 
                   "Language"] <- 'Yukagir'
phonology_coverage[phonology_coverage$Language == 'Tungusic/Even', "Language"] <- 'Even'
phonology_coverage[phonology_coverage$Language == 'Buriat', "Language"] <- 'Buryat'
phonology_coverage[phonology_coverage$Language == 'Uralic/Nganasan', 
                   "Language"] <- 'Nganasan'
phonology_coverage[phonology_coverage$Language == 'Chukchi-Kamchatkan/Koryak', 
                   "Language"] <- 'Koryak'

# Visualize the coverage in a table
inner_join(grammar_coverage, phonology_coverage, by='Language') %>% 
  arrange(desc(Coverage.x)) %>%
kable (booktabs=T, linesep = "", 
      caption ='Language data coverage for all fourteen societies.',
      col.names=c('Language','Grammar (%)', 'Phonology (%)')) %>%
      kable_styling()
```

\clearpage

```{r flatdata, cache=T}

flatten <- function(data.list, gg=siberia_metadata) {
 #' Flattens the nested linguistic data
 #' @param data.list: the input data
 #' @param gg: a data.frame comprising the languages for which the data are flattened
 #' @return the flattened data
  
  df.list <- lapply(seq_along(data.list), function(v) {
    df <- data.list[[v]]
    var.name <- gsub('(.*\\$)', '\\2', names(data.list)[v])
    names(df)[1] <- var.name
    return(dplyr::select(df, UULID, dplyr::everything()))
    })
  df.flat <- Reduce(function(x,y) dplyr::full_join(x, y, by='UULID'), df.list)
  rownames(df.flat) <- sapply(df.flat$UULID, function(x) rownames(gg[gg$UULID %in% x,]))
  return(df.flat)}

# Flatten the data
grammar <- flatten(siberia_grammar_list)
phonology <- flatten(siberia_phonology_list)
```

```{r write to file, include=FALSE, eval=F}
write.csv(grammar, "../data/grammar/grammar_var.csv", na="")
write.csv(phonology, "../data/phonology/phon_var.csv", na="")

```
